---
title: "Are Conservative Cities Smaller in Functional Breadth?"
format:
  clean-revealjs:
    self-contained: true
    embed-resources: true
author:
  - name: Christopher B. Goodman
    orcid: 0000-0002-8565-9714
    email: cgoodman@niu.edu
    affiliations: Northern Illinois University
date: last-modified
bibliography: references.bib
citations-hover: true
#suppress-bibliography: true
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(tibble)
library(pander)
library(scales)
library(haven)
library(readxl)
library(gt)
library(gtsummary)
library(rstatix)
library(fixest)
library("modelsummary")
options(modelsummary_factory_default = "gt")
```

## {.center}

Are cities with residents that hold more ideologically conservative policy positions smaller in functional breadth? 

Do these cities provide fewer services? 

## Motivations

We expect the policy preferences of an area to be reflected in the policies adopted by their government

- Broad support in the political science literature that this happens at the federal and state levels

Does this translation hold for local governments?

- A general assumption that there is no political way to deliver local public services
- Institutional innovations to limit political interference (city managers)

## Motivations

Prior research suggests [@tausanovitchwarshaw2014] that cities with more conservative policy preferences

- Spend less 
- Raise less tax revenue
- Rely on the sales tax more heavily

This analysis examines the first point in more detail. [What is the source of lower spending?]{.alert}

- Less intensity of service provision?
- Fewer services provided?


## Sample Construction {#sec-data}

Period: 2010 (+/-)

Unit: City

- $N = 1,230$ cities

Data: [[Appendix]{.button}](#sec-appendix-data)

- Census of Governments
- Policy conservatism [@tausanovitchwarshaw2014]
- American Community Survey

## Measuring Functional Breadth

:::{.callout-note}
## Functional Performance Index

$$
FP_i = \sum_{n=1}^k f_k \times \frac{E_{jk}}{N_j}
$$ 

where $f_{k}$ is an indicator $[0, 1]$ that municipality $i$ provides service $k$. For each service $k$ a municipality provides, it is weighted by the sample per capita expenditure in category $k$.
:::

:::{.callout-note}
## Functional Inclusiveness or Scope

$$
F_i = \sum_{n=1}^k f_{k}
$$

where $f_k$ is an indicator $[0, 1]$ that municipality $i$ provides service $k$
:::

## {background-image="figures/services.svg" background-size="100%"}

## Estimation Strategy

:::{.callout-note}
## Regression Specification

The main specification is as follows:

$$
\text{Functional Breadth}_i = \alpha +  \text{Citizen Ideology}_i \delta +  X_i \beta + \varepsilon_i
$$
:::

$X_i$ is a vector of variables that control for 

* the size of the city
* the wealth of the city
* the racial composition of the city
* the age of the city

## {background-image="figures/mrp_est.svg" background-size="100%"}

## Descriptive Statistics 


```{r descriptives, include=FALSE}
source("~/Library/CloudStorage/Dropbox/Projects/Papers/FPI-Ideology/analysis/scripts/1_data_import.R")
source("~/Library/CloudStorage/Dropbox/Projects/Papers/FPI-Ideology/analysis/scripts/2_variable_creation.R")

analysis_vars = 
  analysis_vars %>%
  rename(
    "Functional performance index" = fpi_mean,
    "Functions" = functions,
    "Policy conservatism" = mrp_estimate,
    "Population (1000s)" = pop1000,
    "Median household income" = mhi_def,
    "Median house price" = mhp_def,
    "Race index" = d_race,
    "Municipal age" = muniage
  )

```

```{r show-descriptives, echo=FALSE, results="asis"}
analysis_vars %>%
select("Functional performance index", "Functions", "Policy conservatism", "Population (1000s)", "Median household income", "Median house price", "Race index", "Municipal age") %>%
get_summary_stats(show = c("mean", "sd", "min", "max")) %>% 
  gt() %>% 
  cols_label(
    variable = md("Variable"),
    n = md("N"),
    mean = md("Mean"),
    sd = md("SD"),
    min = md("Min"),
    max = md("Max"),
  ) %>%  
  cols_align(
  align = "center", 
  columns = c(n, mean, sd, min, max)
  ) %>%
  cols_align(
  align = "left", 
  columns = variable
  ) %>%
  fmt_number(
    columns = c(mean, sd, min, max),
    decimals = 3,
    use_seps = TRUE
  ) %>%
  opt_align_table_header(align = "center") %>%
  tab_options(
        table_body.hlines.style = "none", 
        column_labels.font.weight = "bold",
        data_row.padding = px(4),
        column_labels.font.size = px(28),
        table.font.size = px(24)
        )

```

## Preliminary Findings

```{r prelim-results, include=FALSE}
source("~/Library/CloudStorage/Dropbox/Projects/Papers/FPI-Ideology/analysis/scripts/2_variable_creation.R")

fpi = feols(
    fpi_mean ~ mrp_estimate + csw0(pop1000 + mhi_def + mhp_def + d_race + muniage + muniage2) | State, data = analysis_vars, vcov = cluster ~ State
)

functions = feols(
    functions ~ mrp_estimate + csw0(pop1000 + mhi_def + mhp_def + d_race + muniage + muniage2) | State, data = analysis_vars, vcov = cluster ~ State
)

dict = c("fpi_mean"             = "Functional performance index",
         "functions"            = "Functions",
         "mrp_estimate"         = "Policy conservatism",
         "pop1000"              = "Population (1000s)",
         "mhi_def"              = "Median household income",
         "mhp_def"              = "Median house price",
         "d_race"               = "Race index",
         "muniage"              = "Municipality age",
         "muniage2"             = "Municipality age squared")

fpi = setNames(fpi, c("(1)", "(2)"))
functions = setNames(functions, c("(3)", "(4)"))

gm <- tibble::tribble(
  ~raw,        ~clean,          ~fmt,
  "nobs",      "N",             0,
  "vcov.type", "SE Clustered", 2)
```


```{r show-prelim-results, echo=FALSE, results="asis"}
modelsummary(
    c(fpi, functions),
    coef_map = dict, stars = TRUE,
    gof_omit = "Adj|IC|F|Log|RMSE|R2",
    gof_map = gm
    ) %>%
    gt::tab_spanner(
        label = "Functional Performance Index",
        columns = 2:3
    ) %>%
    gt::tab_spanner(
        label = "Functions",
        columns = 4:5
    ) %>%
    tab_options(
        table_body.hlines.style = "none", 
        column_labels.font.weight = "bold",
        data_row.padding = px(4),
        column_labels.font.size = px(22),
        table.font.size = px(18)
        )
```

## Extensions

### What about elected mayors or city managers?

```{r extension-results, include=FALSE}
fpi_elected = feols(
    fpi_mean ~ mrp_estimate + csw0(electedmayor + mrp_electedmayor + pop1000 + mhi_def + mhp_def + d_race + muniage + muniage2) | State, data = analysis_vars, vcov = cluster ~ State
)

functions_elected = feols(
    functions ~ mrp_estimate + csw0(electedmayor + mrp_electedmayor + pop1000 + mhi_def + mhp_def + d_race + muniage + muniage2) | State, data = analysis_vars, vcov = cluster ~ State
)

dict = c("fpi_mean"             = "Functional perfomance index",
         "functions"            = "Functions",
         "mrp_estimate"         = "Policy conservatism",
         "electedmayor"         = "Elected mayor (=1)",
         "mrp_electedmayor"     = "Policy conservatism x elected mayor")

fpi_elected = setNames(fpi_elected, c("(1)", "(2)"))
functions_elected = setNames(functions_elected, c("(3)", "(4)"))

gm <- tibble::tribble(
  ~raw,        ~clean,          ~fmt,
  "nobs",      "N",             0,
  "vcov.type", "SE Clustered", 2)
```
```{r show-extension-results, echo=FALSE, results="asis"}
modelsummary(
    c(fpi_elected, functions_elected),
    coef_map = dict, stars = TRUE,
    gof_omit = "Adj|IC|F|Log|RMSE|R2",
    gof_map = gm
    ) %>%
    gt::tab_spanner(
        label = "Functional Performance Index",
        columns = 2:3
    ) %>%
    gt::tab_spanner(
        label = "Functions",
        columns = 4:5
    ) %>%
    tab_options(
        table_body.hlines.style = "none", 
        column_labels.font.weight = "bold",
        data_row.padding = px(4),
        column_labels.font.size = px(22),
        table.font.size = px(18)
        )
```

## Discussion

* Conservative cities are smaller in functional breadth

  * This is largely driven by fewer functions provided
  * Approximately 0.85 to 1.25 fewer functions provided

* Elected mayors eliminate or reduce this effect, suggesting electoral incentives to "do something" may reduce responsiveness


## Next steps

* Explore additional institutional factors that could limit responsiveness
  
  * Partisan council elections, districting, referendum requirements, etc.

* Look at only commonly provided services to

## References

::: {#refs}
:::

## Appendix {#sec-appendix-data visibility="uncounted" .smaller}
### Data

```{r data-sources, include=FALSE}
data_sources <- tribble(
  ~Variable,  ~Source,  ~Year,
  "Functional breadth", "Census of Governments", "2012",
  "Policy conservatism", "@tausanovitchwarshaw2014", "Various",
  "Population", "ACS", "2006-2010",
  "Median household income", "ACS", "2006-2010",
  "Median house price", "ACS", "2006-2010",
  "Race index", "ACS", "2006-2010",
  "Municipal age", "Various", NA
) 
```

```{r show-data-sources, echo=FALSE, results="asis"}
data_sources %>%
  select(Variable, Source, Year) %>% 
  pandoc.table(justify = "lcc", missing = "—")
```


[[Back to main]{.button}](#sec-data)


## Appendix {#sec-appendix visibility="uncounted" .smaller}
```{r base-regressions, include=FALSE}
base_exp = feols(
    pcdirexp ~ mrp_estimate + csw0(pop1000 + mhi_def + mhp_def + d_race + muniage + muniage2) | State, data = analysis_vars, vcov = cluster ~ State
)

base_rev = feols(
    pcrev ~ mrp_estimate + csw0(pop1000 + mhi_def + mhp_def + d_race + muniage + muniage2) | State, data = analysis_vars, vcov = cluster ~ State
)

base_sales = feols(
    share_sales_tax ~ mrp_estimate + csw0(pop1000 + mhi_def + mhp_def + d_race + muniage + muniage2) | State, data = analysis_vars, vcov = cluster ~ State
)

dict = c("pcdirexp"             = "Per capita direct expenditures",
         "pcrev"                = "Per capita total revenue",
         "share_sales_tax"      = "Sales tax share",
         "mrp_estimate"         = "Policy conservatism",
         "pop1000"              = "Population (1000s)",
         "mhi_def"              = "Median household income",
         "mhp_def"              = "Median house price",
         "d_race"               = "Race index",
         "muniage"              = "Municipality age",
         "muniage2"             = "Municipality age squared")


base_exp = setNames(base_exp, c("(1)", "(2)"))
base_rev = setNames(base_rev, c("(3)", "(4)"))
base_sales = setNames(base_sales, c("(5)", "(6)"))
```
```{r show-base-regressions, echo=FALSE, results="asis"}
modelsummary(
    c(base_exp, base_rev, base_sales),
    coef_map = dict, stars = TRUE,
    gof_omit = "Adj|IC|F|Log|RMSE|R2"
    ) %>%
    gt::tab_spanner(
        label = "Per Capita Expenditures",
        columns = 2:3
    ) %>%
    gt::tab_spanner(
        label = "Per Capita Revenues",
        columns = 4:5
    ) %>%
    gt::tab_spanner(
        label = "Sales Tax Share",
        columns = 6:7
  )%>%
  tab_options(
      table_body.hlines.style = "none", 
      column_labels.font.weight = "bold",
      data_row.padding = px(4),
      column_labels.font.size = px(20),
      table.font.size = px(16)
      )
```

[[Back to main]{.button}](#sec-crossref)